---
title: "Assignment 5A - Flight Data"
author: "Carol Campbell"
date: "October 4, 2026"
format: html
editor: visual
---

## Approach

We are tasked with recreating a .csv file provided for us in the assignment. We are free to use any software, including SQL of our choosing, and read it into R for analysis and data manipulation in order to determine which airline has better on-time performance. Is there one carrier who is more reliable than the other and does their location matter?

My plan is to recreate the table in Excel, save it as a .csv file and load it in my Github repository. From there, I will import the file into R using the read_csv function into r. Next I will tidy the data, including pivoting from wide to long format and other exploratory data analysis (EDA) to generate sufficient data that support my final answer.

## Create the data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)
library(ggplot2)
library(kableExtra)
```

## Read file from Github

Using "read_csv" I imported the raw file from my github repository.

```{r read file from Github}
airline_delay = (read_csv("https://raw.githubusercontent.com/carolc57/DATA607/refs/heads/main/cc_airlinedelays.csv"))

airline_delay
 
str(airline_delay)
```

## Remove empty rows; fill empty cells
After importing the file, I see that row three is entirely empty, so I will remove it. Afterwards I will use the fill function to carry the airline names downward to fill the empty N/A cells with the same data. 
```{r missing values check}

#install.packages("janitor")
library(janitor)

airline_delay_mod <- remove_empty(airline_delay, "rows")

airline_delay_mod <- airline_delay_mod|>
  fill(airline, .direction = "down")


#Capitalize the first letter of the first two columns
airline_delay_mod <- airline_delay_mod |> 
  rename_with(str_to_title, .cols = 1:2)

kable(airline_delay_mod, caption = "Tidy Airline Delay")

```

## Pivot to long data frame
Now that my data is "tidy" with missing values replaced, I will now pivot it to a long format to convert the city columns into two variables, Destinations and Flights. Transforming the data into this manner facilitates grouping and analying the dataset.

```{r pivot long}
airline_delay_long <- airline_delay_mod |>
  pivot_longer(
    cols = -c(Airline, Status),
    names_to = "Destination",
    values_to = "Flights"
  ) |>
  mutate(
    Airline = recode(
  Airline,
  "ALASKA" = "Alaska",
  "AM WEST" = "AM West"
),
    Status = str_to_lower(Status)
  )

head(airline_delay_long, 20)


```

## Compare count and percentage of on-time flights for the two airlines.

```{r}
overall_performance <- airline_delay_long |>
  group_by(Airline, Status) |>
  summarise(
    Flights = sum(Flights),
    .groups = "drop"
  ) |>
  group_by(Airline) |>
  mutate(
    Total_flights = sum(Flights),
    Percentage = round((Flights / Total_flights * 100),1)
  )

kable(overall_performance, caption = "Overall Performance")

```
Here we see that AM WEST has the lowest overall delay rate at 10.9% compared to Alaska's 13.3%. Whereas the two airlines are pretty close in their respective on-time rates with AM West leading at 89.1% compared to Alaska's 86.7%


##Let's focus on on-time arrival rates between two by using the filter function
```{r}

overall_performance %>%
  filter(Status == "on time") %>%
  ggplot(aes(x = Airline, y = Percentage, fill = Airline)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(
    aes(label = paste0(round(Percentage, 1), "%")),
    vjust = -0.4
  ) +
  labs(
    title = "Overall Percentage of On-Time Flights",
    x = NULL,
    y = "On Time (%)"
  ) +
  theme_classic()
```
Overall AM West had a slightly better on-time arrival rate than Alaska, however this comparison may be somewhat skewed. Other factors such as destination may tell a different story.

## Compare count and percentage of on-time flights for the two airlines across five cities.

```{r}

city_ontime <- airline_delay_long |>
  group_by(Airline, Destination, Status) |>
  summarise(Flights = sum(Flights), .groups = "drop") |> 
  group_by(Airline, Destination) |> 
  mutate(Total_flights = sum(Flights)) |>                 
  mutate(Percentage = round((Flights / Total_flights * 100), 1)) |>
  filter(Status == "on time") |>
  select(Airline, Destination, Ontime_Flights = Flights,
         Total_flights, OnTime_Percentage = Percentage)  
  
  
kable(city_ontime, caption ="City On-time Performance")
  


```
## On time by City chart
```{r}
ggplot(
  city_ontime,
  aes(
    x = Destination,
    y = OnTime_Percentage,
    fill = Airline
  )
) +
  geom_col(
    position = "dodge"
  ) +
  labs(
    title = "On-Time Percentage by Destination and Airline",
    x = "Destination",
    y = "On-Time Flights (%)",
    fill = "Airline"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )
```
## Key takeaways on City specific On-time performance

1.This graph indicates that Alaska outperforms AM West across the board
For each city—Los Angeles, Phoenix, San Diego, San Francisco, and Seattle—Alaska’s bar is higher. Even without exact numbers, the visual gap is obvious.

2.The performance gap varies by destination Largest gap: Likely San Francisco or Seattle, where Alaska’s bar is much higher.

  Smallest gap: Possibly Phoenix, where the difference looks narrower.

  This suggests that AM West struggles more in certain markets than others.

3. Both airlines show destination‑specific variability
Some cities have noticeably higher on‑time percentages for both airlines (e.g., Seattle), while others show lower performance (e.g., San Francisco).

4. Alaska’s operational consistency stands out
The red bars (Alaska) stay relatively high and stable across destinations. AM West’s teal bars fluctuate more and generally sit lower.










